Varying paths to building the social state in the progressive era

### 2023-11-03 t1_analysis.Rmd
a ddsharecropping and urban immig pop data

@@

```{r read, cache = T}
# ### on libra, read raw
# ### read employment data
# ### full ts of occupations ADD CONSTABLES, THERES A DECENT NUMBER OF THEM
# dat_occ<-read_csv("./data/usa_00071.csv") %>% 
#   rename_all(tolower) %>% 
#   mutate(occ1950 = 
#            case_when(
#              occ1950 == 79 ~ "socwork",
#              occ1950 == 773 ~ "police",
#              occ1950 == 771 ~ "police",
#              occ1950 == 782 ~ "police",
#              occ1950 == 93 ~ "teachers"
#            )) %>% 
#   group_by(year, statefip, occ1950) %>% 
#   summarize(emp = sum(perwt)) %>% 
#   pivot_wider(names_from = occ1950, values_from = emp,
#               values_fill = 0)
# ### AK OR HI HERE? check that at IPUMS
# 
# # pop data (total, age, race, nativitiy, urban)
# dat_immig<-read_csv("./data/usa_00070.csv") %>% 
#   rename_all(tolower) %>% 
#   group_by(year, statefip) %>% 
#   summarize(tot_pop = sum(perwt),
#             children = sum((age<18) * perwt),
#             white_immig = 
#               sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
#             blk = 
#               sum((race == 2) * perwt),
#             urban = sum(hhwt * (urban ==2)),
#             tot_hh = sum(hhwt)) 
# ### AK OR HI HERE? check that at IPUMS
# 
# 
# ### group quarters data, 100% 
# gq<-read_csv("./data/usa_00069.csv") %>% 
#   rename_all(tolower) %>% 
#   filter(age<18) %>% 
#   group_by(year, statefip, gqtype) %>% 
#   summarize(gq_pop = sum(perwt, na.rm = T)) %>% 
#   mutate(gqtype = case_when(
#     gqtype == 2 ~ "child_inst_correctional",
#     gqtype == 3 ~ "child_inst_mental",
#     gqtype == 4 ~ "child_inst_poor_disabled")) %>% 
#   filter(gqtype!="other") %>% 
#   pivot_wider(names_from = gqtype, values_from = gq_pop,
#               values_fill = 0)
# ### AK OR HI HERE? check that at IPUMS
# 
#  
# ### join
# dat <- data.frame(state = state.abb, 
#                region = state.region, 
#                division = state.division) %>% 
#   left_join(state.fips %>% 
#               select(fips, abb) %>% 
#               distinct() %>% 
#               rename(state = abb, statefip = fips)) %>% 
#   left_join(dat_immig) %>% 
#   left_join(dat_occ) %>% 
#   ungroup() %>% 
#   left_join(gq)
# 
# ### homestead act data
# #### load in homestead data and boarding school data
# ### state measure can be cumulative sum of homesteads, that gets at polity
# ### homesteading data
# y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
#   left_join(dat %>% 
#               select(state, region) %>% 
#               distinct()) 
# 
# hs<-read_csv("./data/blm_homesteads_clean.csv") %>% 
#   pivot_longer(claims1868:acres1961,
#                names_to = "name",
#                values_to = "value") %>% 
#   mutate(year = str_sub(
#     name, start = -4, end = -1),
#     type = str_sub(
#       name, start = 1, end = -5)) %>% 
#   select(-name) %>% 
#   mutate(year = as.numeric(year)) %>% 
#   mutate(value = ifelse(is.na(value),
#                         0,
#                         value))
# 
# ### split Dakota Territory values to 0.5 ND, 0.5 SD to preserve time series
# hs_dakota<-hs %>% 
#   filter(state == "Dakota Territory") %>% 
#   mutate(value = value / 2) %>% 
#   select(-state)
# 
# hs<-hs %>% 
#   bind_rows(hs_dakota %>% 
#               mutate(state = "North Dakota")) %>% 
#   bind_rows(hs_dakota %>% 
#               mutate( state = "South Dakota")) %>% 
#   filter(state != "Dakota Territory") %>% 
#   group_by(state, year, type) %>% 
#   summarize(value = sum(value))
#   
# hs<-hs %>% 
#   arrange(state, year) %>% 
#   group_by(state, type) %>% 
#   mutate(cum_value = cumsum(value)) %>% 
#   ungroup()
# 
# ## expand_grid to add zeroes
# ### use y_o_s for index
# state<-unique(y_o_s$statename)
# year<-unique(hs$year)
# type<-unique(hs$type)
# temp<-expand_grid(state, year, type)
# 
# hs<-temp %>% 
#   left_join(hs) %>% 
#   replace_na(list(value = 0,
#                   cum_value = 0))
# 
# hs <- hs %>% 
#   select(-value) %>% 
#   pivot_wider(names_from = type,
#               values_from = cum_value,
#               names_prefix = "hs_") 
# 
# ### add ID and pop vars back to hs
# 
# y_o_s <- read_csv("./data/year_of_statehood.csv") %>% 
#   left_join(dat %>% 
#               select(state, region) %>% 
#               distinct()) %>% 
#   filter(!(is.na(region)))
# 
# hs<-hs %>% 
#   rename(statename = state) %>% 
#   left_join(y_o_s)
# 
# #### join cumulative acres / claims onto dat for the valid census years, then carried forward
# 
# dat<-dat %>% 
#   left_join(hs)

# on local read temp
dat<-read_csv("./data/temp_data.csv")
hs<-read_csv("./data/temp_hs.csv")
```

@@

@@

```{r urban-ts, fig.cap = "Urban population by state 1870-1940"}
p1<-ggplot(dat%>% 
  filter(!(is.na(region))),
       aes(x = year, y = urban / tot_hh,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  labs(subtitle = "Proportion of households urban",
       y = "Proportion",
       x = "Year") + 
  facet_wrap(~region)

p2<-ggplot(dat%>% 
  filter(!(is.na(region))),
       aes(x = year, y = urban,
           group = state)) + 
  geom_line(alpha = 0.5) + 
  labs(subtitle = "Number of households urban",
       y = "Count",
       x = "Year")+ 
  facet_wrap(~region)

>>>>

SETTLER COLONIALISM AS STRUCTURE - PROCESS. BUT SPATIALLY VARIABLE DEPENDING ON LOCAL INTERESTS

@@

```{r read}
# ### on libra, read raw
# ### read employment data
# ### full ts of occupations ADD CONSTABLES, THERES A DECENT NUMBER OF THEM
dat_occ<-read_csv("./data/usa_00071.csv") %>%
  rename_all(tolower) %>%
  mutate(occ1950 =
           case_when(
             occ1950 == 79 ~ "socwork",
             occ1950 == 773 ~ "police",
             occ1950 == 771 ~ "police",
             occ1950 == 782 ~ "police",
             occ1950 == 93 ~ "teachers"
           )) %>%
  group_by(year, statefip, occ1950) %>%
  summarize(emp = sum(perwt)) %>%
  pivot_wider(names_from = occ1950, values_from = emp,
              values_fill = 0)
### AK OR HI HERE? check that at IPUMS

# pop data (total, age, race, nativitiy, urban, farm, landown)
dat_immig<-read_csv("./data/usa_00072.csv") %>%
  rename_all(tolower) %>%
  group_by(year, statefip) %>%
  summarize(tot_pop = sum(perwt),
            children = sum((age<18) * perwt),
            white_immig =
              sum((nativity == 5) * (race == 1) * (hispan == 0) * perwt),
            blk =
              sum((race == 2) * perwt),
            blk_sharecrop = 
              sum((race==2) * (farm==2) * (ownershp==2) * perwt),
            blk_farms = sum((race==2) * (farm==2) * perwt),
            urban_pop = sum(hhwt * (urban ==2)),
            urban_immig_pop = 
              sum((urban==2) * 
                    (nativity == 5) * (race == 1) * (hispan == 0) * perwt),
            tot_hh = sum(hhwt))
### AK OR HI HERE? check that at IPUMS

### group quarters data, 100%
gq<-read_csv("./data/usa_00069.csv") %>%
  rename_all(tolower) %>%
  filter(age<18) %>%
  group_by(year, statefip, gqtype) %>%
  summarize(gq_pop = sum(perwt, na.rm = T)) %>%
  mutate(gqtype = case_when(
    gqtype == 2 ~ "child_inst_correctional",
    gqtype == 3 ~ "child_inst_mental",
    gqtype == 4 ~ "child_inst_poor_disabled")) %>%
  filter(gqtype!="other") %>%
  pivot_wider(names_from = gqtype, values_from = gq_pop,
              values_fill = 0)
### AK OR HI HERE? check that at IPUMS

### join
dat <- data.frame(state = state.abb,
               region = state.region,
               division = state.division) %>%
  left_join(state.fips %>%
              select(fips, abb) %>%
              distinct() %>%
              rename(state = abb, statefip = fips)) %>%
  left_join(dat_immig) %>%
  left_join(dat_occ) %>%
  ungroup() %>%
  left_join(gq)

### homestead act data
#### load in homestead data and boarding school data
### state measure can be cumulative sum of homesteads, that gets at polity
### homesteading data
y_o_s <- read_csv("./data/year_of_statehood.csv") %>%
  left_join(dat %>%
              select(state, region) %>%
              distinct())

hs<-read_csv("./data/blm_homesteads_clean.csv") %>%
  pivot_longer(claims1868:acres1961,
               names_to = "name",
               values_to = "value") %>%
  mutate(year = str_sub(
    name, start = -4, end = -1),
    type = str_sub(
      name, start = 1, end = -5)) %>%
  select(-name) %>%
  mutate(year = as.numeric(year)) %>%
  mutate(value = ifelse(is.na(value),
                        0,
                        value))

### split Dakota Territory values to 0.5 ND, 0.5 SD to preserve time series
hs_dakota<-hs %>%
  filter(state == "Dakota Territory") %>%
  mutate(value = value / 2) %>%
  select(-state)

hs<-hs %>%
  bind_rows(hs_dakota %>%
              mutate(state = "North Dakota")) %>%
  bind_rows(hs_dakota %>%
              mutate( state = "South Dakota")) %>%
  filter(state != "Dakota Territory") %>%
  group_by(state, year, type) %>%
  summarize(value = sum(value))

hs<-hs %>%
  arrange(state, year) %>%
  group_by(state, type) %>%
  mutate(cum_value = cumsum(value)) %>%
  ungroup()

## expand_grid to add zeroes
### use y_o_s for index
state<-unique(y_o_s$statename)
year<-unique(hs$year)
type<-unique(hs$type)
temp<-expand_grid(state, year, type)

hs<-temp %>%
  left_join(hs) %>%
  replace_na(list(value = 0,
                  cum_value = 0))

hs <- hs %>%
  select(-value) %>%
  pivot_wider(names_from = type,
              values_from = cum_value,
              names_prefix = "hs_")

### add ID and pop vars back to hs

y_o_s <- read_csv("./data/year_of_statehood.csv") %>%
  left_join(dat %>%
              select(state, region) %>%
              distinct()) %>%
  filter(!(is.na(region)))

hs<-hs %>%
  rename(statename = state) %>%
  left_join(y_o_s)

#### join cumulative acres / claims onto dat for the valid census years, then carried forward

dat<-dat %>%
  left_join(hs)

# # on local read temp
# dat<-read_csv("./data/temp_data.csv")
# hs<-read_csv("./data/temp_hs.csv")
```

@@

next let's look at sharecropping

```{r}
ggplot(dat %>% 
             filter(year == 1920,
                    blk_farms >= 1e4),
       aes(x = blk_farms, y = blk_sharecrop,
           label = state)) + 
  geom_text(alpha = 0.5) + 
  geom_abline(slope = 1) + 
  labs(x = "Black people living on farms",
       y = "Black people living on tenant farms",
       subtitle = "States with more than 10,000 Black persons living on farms, 1920")
```

And the bivar with sharecropping / blk pop comp

```{r}
ggplot(dat %>% 
         filter(year == 1920),
       aes(x = blk / tot_pop, y = blk_sharecrop,
           label = state,
           color = region)) + 
  #geom_point(alpha = 0.5) + 
  geom_text(alpha = 0.5) + 
  scale_y_log10() + 
  labs(x = "Proportion Black pop",
       y = "Black people living on tenant farms",
       subtitle = "States with more than 10,000 Black persons living on farms, 1920")
```

Places with large Black pops in 1920 also had very high sharecropping rates. Not a perfect correlation, VA and FL are negative off-trend, and lots of variance for low Black pop states, probably just measurement error though. Likely makese sense to use sharecropping as log count rather than as proportion.

@@

```{r urban-ts, fig.cap = "Urban white immigrant population by state 1870-1940"}
p1<-ggplot(dat%>% 
             filter(year<=1930) %>% 
             filter(!(is.na(region))),
           aes(x = year, y = urban_immig_pop / tot_pop,
               group = state)) + 
  geom_line(alpha = 0.5) + 
  labs(subtitle = "Proportion of population white immigrant urban",
       y = "Proportion",
       x = "Year") + 
  facet_wrap(~region)

p2<-ggplot(dat%>% 
             filter(year<=1930) %>% 
             filter(!(is.na(region))),
           aes(x = year, y = urban_immig_pop,
               group = state)) + 
  geom_line(alpha = 0.5) + 
  labs(subtitle = "White immigrant urban population",
       y = "Count",
       x = "Year")+ 
  facet_wrap(~region)
